---
title: Sweet Mary Momentum
author: Fennon W.
date: '2019-12-01'
slug: momentum
categories:
  - Finance
tags:
  - algotrading
  - r
  - alpha vantage
  - financial research
disableComments: no
---

Can a <a href="https://www.investopedia.com/terms/m/momentum.asp">momentum</a> strategy beat the gold standard Buy And Hold? According to some talented <a href="http://www.business.unr.edu/faculty/liuc/files/BADM742/Jegadeesh_Titman_1993.pdf">financial researchers</a> and <a href="https://www.aqr.com/-/media/AQR/Documents/Insights/White-Papers/The-Case-for-Momentum-Investing.pdf">asset managers</a>, the answer is yes. 

Let's see for ourselves. We'll use a long-only strategy and trade **SPY**, the SPDR S&P500 ETF designed to track the S&P 500 stock market index. Momentum will be measured based on monthly returns over the prior 3, 6, 9, and 12 months.

Historical ETF data covers from 1999-2019. Feel free to explore the data yourself using the <a href="https://github.com/business-science/alphavantager">alphavantager</a> R package and <a href="https://www.alphavantage.co/documentation">Alpha Vantage API</a>.


# Setup

Load libraries.

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(alphavantager)
library(scales)
library(knitr)
library(kableExtra)
```

```
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(alphavantager)
library(scales)
library(knitr)
```

Set Alphavantager key.
```{r eval = FALSE}
av_api_key("my_key_goes_here")
```

```{r echo = FALSE}
av_api_key("ZVWHBIOOHGCNKHD3")
```

# Get Historical ETF Data

Get all available price data on a monthly timeframe.
```{r echo = FALSE, eval = FALSE, message = FALSE}
av_get(symbol = "SPY"
       , av_fun = "TIME_SERIES_MONTHLY_ADJUSTED"
       , outputsize = "full") -> all_mthly
```

```{r echo = FALSE}
all_mthly <- readRDS("/Users/Fennon/Desktop/My Blog/projects/*DATA*/all_mthly_spy.rds")
```


Calculate momentum as the return over the past n = 3, 6, 9, and 12 months based on month close. We'll get the past n months return by using lag(), but since this current month's close counts, we'll use lag(n-1).
```{r warning = FALSE}
all_mthly %>% 
  mutate(m3 = close - lag(close, 2)
         , m6 = close - lag(close, 5)
         , m9 = close - lag(close, 8)
         , m12 = close - lag(close, 11)) -> all_momo

```

```{r}
head(all_momo, 5)
```



# Buy/Sell Rules
According to the research, if momentum (MOMO) is positive, price is likely to continue in the same positive direction. So we'll make the following buy/sell decisions:

* Buy if MOMO > 0 (unless already in a trade)
* Sell if MOMO < 0 (unless already out of a trade)

```{r}
all_momo %>% 
  mutate(signal3 = case_when(m3 > 0 ~ "buy"
                              , m3 < 0 ~ "sell")
         , signal6 = case_when(m6 > 0 ~ "buy"
                               , m6 < 0 ~ "sell")
         , signal9 = case_when(m9 > 0 ~ "buy"
                               , m9 < 0 ~ "sell")
         , signal12 = case_when(m12 > 0 ~ "buy"
                                , m12 < 0 ~ "sell")
         
         , previous_signal3 = lag(signal3, 1)
         , previous_signal6 = lag(signal6, 1)
         , previous_signal9 = lag(signal9, 1)
         , previous_signal12 = lag(signal12, 1)
         
         # only buy and sell when current signal different from previous signal
         , decision3 = case_when(signal3 == previous_signal3 & previous_signal3 == "buy" ~ "hold"
                                 , signal3 == previous_signal3 & previous_signal3 == "sell" ~ "out"
                                  , TRUE ~ signal3)
         , decision6 = case_when(signal6 == previous_signal6 & previous_signal6 == "buy" ~ "hold"
                                 , signal6 == previous_signal6 & previous_signal6 == "sell" ~ "out"
                                 , TRUE ~ signal6)
         , decision9 = case_when(signal9 == previous_signal9 & previous_signal9 == "buy" ~ "hold"
                                 , signal9 == previous_signal9 & previous_signal9 == "sell" ~ "out"
                                 , TRUE ~ signal9)
         , decision12 = case_when(signal12 == previous_signal12 & previous_signal12 == "buy" ~ "hold"
                                  , signal12 == previous_signal12 & previous_signal12 == "sell" ~ "out"
                                  , TRUE ~ signal12)) %>% 
  data.frame() -> momo_trade
```


Below, the trading descisions are plotted for each momentum strategy. Verticals lines represent <span style="color:green">buy</span> and <span style="color:red">sell</span>.
```{r echo = FALSE, fig.width = 8.5, fig.height = 6}

for(i in c("3", "6", "9", "12")){
  
  momo_trade %>% 
    select(timestamp, close, ends_with(i)) %>% 
    rename("decision" = glue::glue("decision{i}")
           , "previous_signal" = glue::glue("previous_signal{i}")) %>% 
    as.data.frame() -> momo_trade_m
  
  momo_trade_m %>% 
    ggplot(aes(x = timestamp, y = close)) +
    geom_line() + 
    scale_x_date(date_breaks = "years" , date_labels = "%Y") +
    geom_vline(data = momo_trade_m %>% filter(decision == "buy"),
               aes(xintercept = timestamp),
               col = "green3",
               linetype = "dashed") +
    geom_vline(data = momo_trade_m %>% filter(decision == "sell"),
               aes(xintercept = timestamp),
               col = "red") +
    xlab("Date") +
    ylab("Close") +
    theme_classic() +
    theme(legend.title = element_blank()) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle(glue::glue("{i}mo")) -> plot_m
  
  assign(glue::glue("plot_m{i}"), plot_m)


}

grid.arrange(plot_m3, plot_m6, plot_m9, plot_m12)

```

While it may not seem obvious in the 3mo and 6mo, the plots generally show momentum helped us weather the financial storms of the tech bubble (2000-2002), housing crisis (2007-2009), and mini-wtf in late 2018. Below are the same plots, but with financial crises shaded in grey.

```{r echo = FALSE, fig.width = 8.5, fig.height = 6}
plot_m3 +
  geom_rect(aes(xmin = as_date("2000-03-11"), xmax = as_date("2002-10-09")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2007-12-01"), xmax = as_date("2009-06-01")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2018-09-20"), xmax = as_date("2018-12-24")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) -> plot_m3_crisis

plot_m6 +
  geom_rect(aes(xmin = as_date("2000-03-11"), xmax = as_date("2002-10-09")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2007-12-01"), xmax = as_date("2009-06-01")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2018-09-20"), xmax = as_date("2018-12-24")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) -> plot_m6_crisis

plot_m9 +
  geom_rect(aes(xmin = as_date("2000-03-11"), xmax = as_date("2002-10-09")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2007-12-01"), xmax = as_date("2009-06-01")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2018-09-20"), xmax = as_date("2018-12-24")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) -> plot_m9_crisis

plot_m12 +
  geom_rect(aes(xmin = as_date("2000-03-11"), xmax = as_date("2002-10-09")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2007-12-01"), xmax = as_date("2009-06-01")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) +
  geom_rect(aes(xmin = as_date("2018-09-20"), xmax = as_date("2018-12-24")
                , ymin = -Inf, ymax = Inf), alpha = 0.009) -> plot_m12_crisis


grid.arrange(plot_m3_crisis, plot_m6_crisis, plot_m9_crisis, plot_m12_crisis)
```

With avoiding those scary times, it looks like these strategies actually could produce better returns with lower drawdowns than Buy and Hold. Let's compare all strategies using the metric RoMad -- return over maximum drawdown. This helps us gauge how much pain we'd be willing to take to capture the gains (if any).

```{r warning = FALSE}
## trade performance vs buy and hold
for(i in c("3", "6", "9", "12")){
  
  if(i == "3"){
    
    # calc monthly returns while in an open trade + holding
    momo_trade %>% 
      select(timestamp, close, ends_with(i)) %>% 
      rename("decision" = glue::glue("decision{i}")
             , "previous_signal" = glue::glue("previous_signal{i}")) %>% 
      filter(!is.na(decision)) %>% 
      mutate(return_tmp = (close - lag(close, 1)) / lag(close, 1)
             , return = ifelse(previous_signal == "buy", return_tmp, 0)
             , return = ifelse(is.na(return), 0 , return)
             , cum_return = cumprod(1 + return)
             , account = 10000 * cum_return
             , momo = as.character(glue::glue("{i}mo"))) %>% 
      select(momo, timestamp, close, previous_signal, decision
             , return, cum_return, account) -> momo_perf
      
  
    }else{
      
      # calc monthly returns while in an open trade + holding
      momo_trade %>% 
        select(timestamp, close, ends_with(i)) %>% 
        rename("decision" = glue::glue("decision{i}")
               , "previous_signal" = glue::glue("previous_signal{i}")) %>% 
        filter(!is.na(decision)) %>% 
        mutate(return_tmp = (close - lag(close, 1)) / lag(close, 1)
               , return = ifelse(previous_signal == "buy", return_tmp, 0)
               , return = ifelse(is.na(return), 0 , return)
               , cum_return = cumprod(1 + return)
               , account = 10000 * cum_return
               , momo = as.character(glue::glue("{i}mo"))) %>% 
        select(momo, timestamp, close, previous_signal, decision
               , return, cum_return, account) -> momo_perf_tmp
      
      momo_perf %>% 
        bind_rows(momo_perf_tmp) -> momo_perf
  }
}

## add in buy and hold perf
all_mthly %>% 
  mutate(return = (close - lag(close, 1)) / lag(close, 1)
         , return = ifelse(is.na(return), 0, return)
         , cum_return = cumprod(1 + return)
         , account = 10000 * cum_return
         , momo = "buy_hold"
         , previous_signal = "hold"
         , decision = "hold") %>% 
  select(momo, timestamp, close, previous_signal, decision
         , return, cum_return, account) %>% 
  data.frame -> buy_hold_perf

momo_perf %>% 
  bind_rows(buy_hold_perf) -> all_perf

all_perf %>% 
  group_by(momo) %>% 
  mutate(port_return_tmp = ifelse(row_number() == n(), cum_return, NA)
         , port_return = port_return_tmp - 1) %>% 
  filter(!is.na(port_return)) %>% 
  select(momo, port_return) %>% 
  inner_join(all_perf, by = c("momo")) %>% 
  group_by(momo, port_return) %>% 
  summarise(max_drawdown = maxDrawdown(return)) %>% 
  ungroup() %>% 
  mutate(romad = port_return / max_drawdown) %>% 
  as.data.frame() -> port_analytics


## kable for a pretty table
port_analytics %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")
                , full_width = FALSE
                , digits = 2)
```

 

Words

```{r echo = FALSE}
## plot momo returns (first point represents first buy signal, all others time of sell)
all_perf$momo <- factor(all_perf$momo, levels = c("3mo", "6mo", "9mo", "12mo", "buy_hold"))

all_perf %>% 
  ggplot(aes(x = timestamp, y = account)) +
  geom_line(aes(color = momo)) +
  # geom_line(data = all_perf %>% filter(momo != "buy_hold"), aes(color = momo)) +
  # geom_line(data = all_perf %>% filter(momo == "buy_hold"), aes(color = momo), color = "black") +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  scale_y_continuous(limits = c(0, 40000)) + 
  xlab("Date") +
  ylab("Account") +
  #labs(color = "Momentum") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.title = element_blank()) +
  ggtitle("Momo vs Buy And Hold") -> plot_returns

plot_returns
```

