---
title: EMA SMA Cross Algorithm & Backtest
author: FW
date: '2019-11-03'
slug: ema-sma-cross-algorithm-backtest
categories: [Fintech]
tags:
  - r
  - algotrading
disableComments: no
---

This algorithm is based on the "moving average crossover" strategy. Traditionally, this strategy involves taking two moving averages -- a "slow" one that averages over a longer time window and a "fast" one that averages over a shorter time window -- and buy/sell based on when the two moving averages crossover.

I used a 50-day simple moving average (SMA) as the "slow" one and 10, 20, 30, and 40-day exponential moving averages (EMA) as the fast ones. For each SMA/EMA pair, I bought when the EMA crossed above the SMA and sold when the EMA crossed below. I decided to trade **SPY**, the SPDR S&P500 ETF. This ETF is very liquid, covers a broad range of stocks (so no cherry picking), and can be traded commission free by most brokerages. **However, trade at your own risk.** Neither I, nor your parents, nor Jesus are responsible for your financial decisions. 

To obtain historical ETF data, I used the <a href="https://github.com/business-science/alphavantager">alphavantager</a> R package to access the <a href="https://www.alphavantage.co/documentation/">Alpha Vantage API</a>; the <a href="https://github.com/business-science/tidyquant">tidyquant</a> package to calculate simple moving averages; and tidyverse, ggplot2, etc to wrangle and visualize data.

The alogirthm was trained and tested on a 80/20 split of all available daily price data through alphavantager at the time of this analysis. This covered approximately 20 years, from 1999 to 2019.

## Assumptions

* Starting investment of $10,000 USD
* Long-only strategy
* Trade decisions based on daily close price, but trade entry/exit based on next day open
* Train set = data from 1999 to 2015
* Test set = data from 2015 to 2019
* Strategy to beat: buy and hold over entire time period

**TL;DR: My algorithm did not make money and I'm still not a billionaire :( **

***

Load libraries.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(lubridate)
library(alphavantager)
library(scales)
```

```
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(lubridate)
library(alphavantager)
library(scales)
```

Set Alphavantager key.
```{r}
# av_api_key("my_key_goes_here"))
```
```{r, echo = FALSE}
av_api_key("ZVWHBIOOHGCNKHD3")
```

# Get historical ETF data

Get all available daily price data and split into train and test sets.

```{r, message = FALSE, warning = FALSE}
# av_get(symbol = "SPY"
#        , av_fun = "TIME_SERIES_DAILY_ADJUSTED"
#        , outputsize = "full") %>%
#   mutate(row_id = row_number()) %>%
#   select(row_id, timestamp, open, close) -> all_dat

# train_dat <- all_dat %>% filter(row_id <= floor(nrow(all_dat)*.80))
# test_dat <- all_dat %>% filter(!row_id %in% train_dat$row_id)
```

```{r, echo = FALSE}
all_dat <- readRDS("/Users/Fennon/Desktop/My Blog/projects/ema_sma_cross/all_dat.rds")
train_dat <- all_dat %>% filter(row_id <= floor(nrow(all_dat)*.80))
test_dat <- all_dat %>% filter(!row_id %in% train_dat$row_id)
```


# Moving Averages

Calculate EMAs and SMA over all desired time windows.

```{r, message = FALSE, warning = FALSE}
train_dat %>%
  mutate(EMA10 = EMA(close, n = 10)
         , EMA20 = EMA(close, n = 20)
         , EMA30 = EMA(close, n = 30)
         , EMA40 = EMA(close, n = 40)
         , SMA50 = SMA(close, n = 50)) %>%
  filter(!is.na(SMA50)) -> train_dat_w_mas
```

Convert data format from wide to long.

```{r, warning = FALSE}
train_dat_w_mas %>%
  select(-row_id) %>% 
  gather("var", "value", 2:8) -> train_dat_w_mas_long
```


Plot daily close price and moving averages.

```{r, warning = FALSE}
train_dat_w_mas_long %>% 
  filter(!var %in% c("open")) %>% 
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(group = factor(var)), size = 0.5) +
  scale_color_manual(values = c("black", "orange", "green", "turquoise", "magenta", "blue")) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("SPY with Moving Averages")
```

Oof, that's a lot of data to take in. To better see the EMA/SMA relationship, here's a closer look at the couple of years.

```{r, echo = FALSE, warning = FALSE}
train_dat_w_mas_long %>% 
  filter(!var %in% c("open")
         , timestamp <= "2002-01-01" ) %>% 
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(group = factor(var)), size = 0.5) +
  scale_color_manual(values = c("black", "orange", "green", "turquoise", "magenta", "blue")) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("SPY with Moving Averages")
```

# Encode buy/sell signals and decisions

Buy/sell decisions are made as follows:
* Buy if EMA > SMA (unless I bought already)
* Sell if EMA < SMA (unless I sold already)

I code the decisions in two passes: the first pass codes the signal, and the second pass codes the decision. Basically, the decision coding will make sure I don't buy again if the last action was to buy, and I won't sell again if the last action was to sell. 


