---
title: EMA x SMA Algorithm & Backtest
author: FW
date: '2019-11-03'
slug: ema-sma-cross-algorithm-backtest
categories: [Finance]
tags:
  - r
  - algotrading
disableComments: no
---

This algorithm is based on the "moving average crossover" strategy. Traditionally, this strategy involves taking two moving averages -- a "slow" one that averages over a longer time window and a "fast" one that averages over a shorter time window -- and buy/sell based on when the two moving averages crossover.

I used a 50-day simple moving average (SMA) as the "slow" one and 10, 20, 30, and 40-day exponential moving averages (EMA) as the fast ones. For each SMA/EMA pair, I bought when the EMA crossed above the SMA and sold when the EMA crossed below. I decided to trade **SPY**, the SPDR S&P500 ETF designed to track the S&P 500 stock market index. This ETF is very liquid, covers a broad range of stocks (so no cherry picking), and can be traded commission free by most brokerages. **However, trade at your own risk.** Neither I, nor your parents, nor Jesus are responsible for your financial decisions. 

To obtain historical ETF data, I used the <a href="https://github.com/business-science/alphavantager">alphavantager</a> R package to access the <a href="https://www.alphavantage.co/documentation/">Alpha Vantage API</a>; the <a href="https://github.com/business-science/tidyquant">tidyquant</a> package to calculate simple moving averages; and tidyverse, ggplot2, etc to wrangle and visualize data.

The algorithm was trained (in-sample) and tested (out-of-sample) on a 80/20 split of all available daily price data through alphavantager at the time of this analysis. This covered approximately 20 years, from 1999 to 2019.

## Assumptions

* Starting investment of $10,000 USD
* Long-only strategy
* No slippage, no commission
* Train Set: data from 1999 to 2015
* Test Set: data from 2015 to 2019
* Strategy to beat: buy and hold over entire time period

**TL;DR: My algorithm did not make money and I'm still poor :( **

***
# Setup

Load libraries.

```{r echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(alphavantager)
library(scales)
```

```
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(lubridate)
library(alphavantager)
library(scales)
```

Set Alphavantager key.
```{r eval = FALSE}
av_api_key("my_key_goes_here")
```

```{r echo = FALSE}
av_api_key("ZVWHBIOOHGCNKHD3")
```

# Get Historical ETF Data

Get all available daily price data and split into train and test sets.

```{r eval = FALSE, message = FALSE}
av_get(symbol = "SPY"
       , av_fun = "TIME_SERIES_DAILY_ADJUSTED"
       , outputsize = "full") %>%
  mutate(row_id = row_number()) %>%
  select(row_id, timestamp, open, close) -> all_dat

all_dat %>% filter(row_id <= floor(nrow(all_dat)*.80)) -> train_dat
all_dat %>% filter(!row_id %in% train_dat$row_id) -> test_dat
```

```{r echo = FALSE}
all_dat <- readRDS("/Users/Fennon/Desktop/My Blog/projects/ema_sma_cross/all_dat.rds")

all_dat %>% filter(row_id <= floor(nrow(all_dat)*.80)) -> train_dat
all_dat %>% filter(!row_id %in% train_dat$row_id) -> test_dat
```

# Moving Averages

Calculate EMAs and SMA over all desired time windows.

```{r message = FALSE, warning = FALSE}
train_dat %>%
  mutate(EMA10 = EMA(close, n = 10)
         , EMA20 = EMA(close, n = 20)
         , EMA30 = EMA(close, n = 30)
         , EMA40 = EMA(close, n = 40)
         , SMA50 = SMA(close, n = 50)) %>%
  filter(!is.na(SMA50)) -> train_dat_w_mas
```

Convert data format from wide to long.

```{r warning = FALSE}
train_dat_w_mas %>%
  select(-row_id) %>% 
  gather("var", "value", 2:8) -> train_dat_w_mas_long
```


Plot daily close price and moving averages.

```{r warning = FALSE}
train_dat_w_mas_long %>% 
  filter(!var %in% c("open")) %>% 
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(group = factor(var)), size = 0.5) +
  scale_color_manual(values = c("black", "orange", "green", "turquoise", "magenta", "blue")) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("SPY with Moving Averages")
```

Wow, what a clear chart...To better see the EMA/SMA relationship, here's a closer look at the first couple of years.

```{r echo = FALSE, warning = FALSE}
train_dat_w_mas_long %>% 
  filter(!var %in% c("open")
         , timestamp <= "2002-01-01" ) %>% 
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(group = factor(var)), size = 0.5) +
  scale_color_manual(values = c("black", "orange", "green", "turquoise", "magenta", "blue")) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("SPY with Moving Averages")
```

# Encode buy/sell signals and decisions

Buy/sell decisions are made as follows:

* Buy if EMA > SMA (unless I bought already)
* Sell if EMA < SMA (unless I sold already)

We'll encode the decisions in two passes for each EMA/SMA crossover algo strategy: the first pass encodes the signal, and the second pass encodes the decision. Basically, the decision encoding will make sure we don't buy again if the last action was to buy, and I don't sell again if the last action was to sell. For each decision, the trade is entered or exited on the next day's open.

```{r}
train_dat_w_mas %>% 
  # Buy/Sell signal
  mutate(signal10 = case_when(EMA10 > SMA50 ~ "buy"
                            , EMA10 < SMA50 ~ "sell")
         , signal20 = case_when(EMA20 > SMA50 ~ "buy"
                              , EMA20 < SMA50 ~ "sell")
         , signal30 = case_when(EMA30 > SMA50 ~ "buy"
                              , EMA30 < SMA50 ~ "sell")
         , signal40 = case_when(EMA40 > SMA50 ~ "buy"
                              , EMA40 < SMA50 ~ "sell")
         
         # Create lagging indicator for action
         , previous_signal10 = lag(signal10, 1)
         , previous_signal20 = lag(signal20, 1)
         , previous_signal30 = lag(signal30, 1)
         , previous_signal40 = lag(signal40, 1)
         
         # Encode hold as any time current and last action are the same
         # since we don't want to keep buying/selling
         , decision10 = case_when(signal10 == previous_signal10 ~ "hold"
                                , TRUE ~ signal10)
         , decision20 = case_when(signal20 == previous_signal20 ~ "hold"
                                , TRUE ~ signal20)
         , decision30 = case_when(signal30 == previous_signal30 ~ "hold"
                                , TRUE ~ signal30)
         , decision40 = case_when(signal40 == previous_signal40 ~ "hold"
                                , TRUE ~ signal40)) -> all_trades
```

Extract the periods when we would have bought or sold. 
```{r}
# EMA10 / SMA50
all_trades %>% 
  filter(decision10 %in% c("buy", "sell")) %>% 
  select(row_id, timestamp, open, close, EMA10, SMA50, decision = decision10) -> trades10

# EMA20 / SMA50
all_trades %>% 
  filter(decision20 %in% c("buy", "sell")) %>% 
  select(row_id, timestamp, open, close, EMA20, SMA50, decision = decision20) -> trades20

# EMA30 / SMA50
all_trades %>% 
  filter(decision30 %in% c("buy", "sell")) %>% 
  select(row_id, timestamp, open, close, EMA30, SMA50, decision = decision30) -> trades30

# EMA40 / SMA50
all_trades %>% 
  filter(decision40 %in% c("buy", "sell")) %>% 
  select(row_id, timestamp, open, close, EMA40, SMA50, decision = decision40) -> trades40
```

Let's take a look at the EMA10/SMA50 trade decisions.
```{r}
head(trades10, 5)
```

# Visualizing Algorithm Decisions

We'll visualize trading decisions by plotting them as vertical intercept lines over the graph (<span style="color:green">green</span> = buy, <span style="color:red">red</span> = sell). The first couple years of trade decisions for each EMA/SMA pair shown below.

```{r echo = FALSE}
train_dat_w_mas_long %>%
  filter(var %in% c("close", "EMA10", "SMA50")) %>% 
  filter(timestamp < "2002-01-01") %>%
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(color = var, size = var)) +
  scale_color_manual(values = c("black", "orange", "blue")) +
  scale_size_manual(values = c(0.5, 0.5, 0.5)) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  geom_vline(data = trades10 %>% filter(decision == "buy", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "green3",
             linetype="dashed") +
  geom_vline(data = trades10 %>% filter(decision == "sell", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "red") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  ggtitle("EMA10") -> plot_ema10

train_dat_w_mas_long %>%
  filter(var %in% c("close", "EMA20", "SMA50")) %>% 
  filter(timestamp < "2002-01-01") %>%
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(color = var, size = var)) +
  scale_color_manual(values = c("black", "green", "blue")) +
  scale_size_manual(values = c(0.5, 0.5, 0.5)) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  geom_vline(data = trades20 %>% filter(decision == "buy", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "green3",
             linetype="dashed") +
  geom_vline(data = trades20 %>% filter(decision == "sell", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "red") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  ggtitle("EMA20") -> plot_ema20

train_dat_w_mas_long %>%
  filter(var %in% c("close", "EMA30", "SMA50")) %>%
  filter(timestamp < "2002-01-01") %>%
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(color = var, size = var)) +
  scale_color_manual(values = c("black", "turquoise", "blue")) +
  scale_size_manual(values = c(0.5, 0.5, 0.5)) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  geom_vline(data = trades30 %>% filter(decision == "buy", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "green3",
             linetype="dashed") +
  geom_vline(data = trades30 %>% filter(decision == "sell", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "red") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  ggtitle("EMA30") -> plot_ema30

train_dat_w_mas_long %>%
  filter(var %in% c("close", "EMA40", "SMA50")) %>%
  filter(timestamp < "2002-01-01") %>% 
  ggplot(aes(x=timestamp, y=value, color = var)) +
  geom_line(aes(color = var, size = var)) +
  scale_color_manual(values = c("black", "magenta", "blue")) +
  scale_size_manual(values = c(0.5, 0.5, 0.5)) +
  scale_x_date(date_breaks = "years" , date_labels = "%Y") +
  xlab("Date") +
  ylab("Close") +
  geom_vline(data = trades40 %>% filter(decision == "buy", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "green3",
             linetype="dashed") +
  geom_vline(data = trades40 %>% filter(decision == "sell", timestamp < "2002-01-01"),
             aes(xintercept = timestamp),
             col = "red") +
  theme_classic() +
  theme(legend.title = element_blank()) +
  ggtitle("EMA40") -> plot_ema40
```

```{r echo = FALSE, fig.width = 8.5, fig.height = 6}
grid.arrange(plot_ema10, plot_ema20, plot_ema30, plot_ema40, nrow = 2)
```

# Simulate Backtest

We'll calculate profit/loss of each EMA/SMA algo. Let's use EMA10/SMA50 trades (trade10) as an example to see how we re-format the data for the backtest. First, we create a dataset of buy decisions, give each trade a trade_id, and join back to the entire train dataset to get the next day's open (our actual trade price). Second, we do the same for sell decisions, giving each trade the same trade_id as its corresponding buy decision. And third, we join the two datasets, with each row representing a roundtrip trade, and are now ready to calculate profit/loss.

```{r echo = FALSE}
# EMA10
trades10 %>% 
  filter(decision == "buy") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>% 
  inner_join(train_dat, by=c("next_id"="row_id")) %>% 
  select(trade_id, row_id, buy_timestamp = timestamp.y, buy_price = open.y) -> trades10_buy

trades10 %>%
  filter(decision == "sell") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>%
  inner_join(train_dat, by=c("next_id"="row_id")) %>%
  select(trade_id, row_id, sell_timestamp = timestamp.y, sell_price = open.y) -> trades10_sell

trades10_buy %>%
  inner_join(trades10_sell, by=c("trade_id" = "trade_id")) -> trades10_ready

# EMA20
trades20 %>% 
  filter(decision == "buy") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>% 
  inner_join(train_dat, by=c("next_id"="row_id")) %>% 
  select(trade_id, row_id, buy_timestamp = timestamp.y, buy_price = open.y) -> trades20_buy

trades20 %>%
  filter(decision == "sell") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>%
  inner_join(train_dat, by=c("next_id"="row_id")) %>%
  select(trade_id, row_id, sell_timestamp = timestamp.y, sell_price = open.y) -> trades20_sell

trades20_buy %>%
  inner_join(trades20_sell, by=c("trade_id" = "trade_id")) -> trades20_ready

# EMA30
trades30 %>% 
  filter(decision == "buy") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>% 
  inner_join(train_dat, by=c("next_id"="row_id")) %>% 
  select(trade_id, row_id, buy_timestamp = timestamp.y, buy_price = open.y) -> trades30_buy

trades30 %>%
  filter(decision == "sell") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>%
  inner_join(train_dat, by=c("next_id"="row_id")) %>%
  select(trade_id, row_id, sell_timestamp = timestamp.y, sell_price = open.y) -> trades30_sell

trades30_buy %>%
  inner_join(trades30_sell, by=c("trade_id" = "trade_id")) -> trades30_ready

# EMA40
trades40 %>% 
  filter(decision == "buy") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>% 
  inner_join(train_dat, by=c("next_id"="row_id")) %>% 
  select(trade_id, row_id, buy_timestamp = timestamp.y, buy_price = open.y) -> trades40_buy

trades40 %>%
  filter(decision == "sell") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>%
  inner_join(train_dat, by=c("next_id"="row_id")) %>%
  select(trade_id, row_id, sell_timestamp = timestamp.y, sell_price = open.y) -> trades40_sell

trades40_buy %>%
  inner_join(trades40_sell, by=c("trade_id" = "trade_id")) -> trades40_ready


trades_ready <- list(trades10_ready, trades20_ready, trades30_ready, trades40_ready)
names(trades_ready) <- c("EMA10", "EMA20", "EMA30", "EMA40")
```


```{r}
trades10 %>% 
  filter(decision == "buy") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>% 
  inner_join(train_dat, by=c("next_id"="row_id")) %>% 
  select(trade_id, buy_timestamp = timestamp.y, buy_price = open.y) -> trades10_buy

trades10 %>%
  filter(decision == "sell") %>% 
  mutate(next_id = row_id + 1
         , trade_id = row_number()) %>%
  inner_join(train_dat, by=c("next_id"="row_id")) %>%
  select(trade_id, sell_timestamp = timestamp.y, sell_price = open.y) -> trades10_sell

trades10_buy %>%
  inner_join(trades10_sell, by=c("trade_id" = "trade_id")) %>% 
  select(trade_id, buy_timestamp, buy_price, sell_timestamp, sell_price) %>% 
  as.data.frame() -> trades10_ready

head(trades10_ready)
```

We repeat this process for each of the other EMA/SMA algos and create a list of datasets of each algo's trades (trade_ready). We then loop through each dataset, loop through all trades within, and calculate profit/loss. If the account value ever reaches <$0 (aka "ACCOUNT BLOWN!"), the backtest stops.

```{r}

# initial investment
account <- 10000

# dump all trades + info here
trade_log <- data.frame(cross_strat = as.character()
                      , trade_n = as.numeric()
                      , shares = as.numeric()
                      , p_l = as.numeric()
                      , capital = as.numeric())

for(i in 1:length(names(trades_ready))){
  
  # grab trades for EMA/SMA crossover algo strategy (cross_strat)
  trades_ready_slim <- trades_ready[[i]]
  cross_strat <- names(trades_ready)[i]
  
  for(j in 1:nrow(trades_ready_slim)){
    
    # first trade
    if(j == 1){
      
      trade_n <- j
      shares <- floor(account/trades_ready_slim$buy_price[j])
      p_l <- (trades_ready_slim$sell_price[j]-trades_ready_slim$buy_price[j])*shares
      capital <- account+p_l
      
      log_entry<-cbind.data.frame(cross_strat, trade_n, shares, p_l, capital)
      
      trade_log %>%
        rbind(log_entry) -> trade_log
      
    }
    
    # all other trades
    if(j > 1){
      
      if(trade_log$capital[j-1]>0){
        
        trade_n <- j
        shares <- floor(as.numeric(trade_log$capital[j-1])/trades_ready_slim$buy_price[j])
        p_l <- (trades_ready_slim$sell_price[j]-trades_ready_slim$buy_price[j])*shares
        capital <- as.numeric(trade_log$capital[j-1])+p_l
        
        log_entry <- cbind.data.frame(cross_strat, trade_n, shares, p_l, capital)
        
        trade_log %>%
          rbind(log_entry) -> trade_log
        
      }
      
      # worst case scenario
      if(trade_log$capital[j-1]<=0){
        stop("ACCOUNT BLOWN!")
        
      }
    }
  }
}

```

```{r}
head(trade_log)
```

# Final Results

With all trades now analyzed, there's only one thing left to do: calculate ROI and compare to benchmark Buy & Hold strategy.

```{r}
trade_log %>% 
  group_by(cross_strat) %>% 
  summarise(max_n = max(trade_n)) %>% 
  ungroup() %>% 
  inner_join(trade_log, by = c("cross_strat")) %>% 
  filter(trade_n == max_n) %>% 
  mutate(ROI = percent((capital - account)/account)) %>% 
  select(cross_strat, ROI) -> algo_roi

# benchmark strat
train_dat %>% 
  filter(row_id == 1) %>% 
  select(-close) -> bh_entry

train_dat %>% 
  filter(row_id == nrow(train_dat)) %>% 
  mutate(row_id = 1) %>% 
  select(-open) -> bh_exit

bh_entry %>% 
  inner_join (bh_exit, by = c("row_id")) %>% 
  mutate(cross_strat = "Buy & Hold"
         , trade_n = 1
         , shares = floor(account/open)
         , p_l = (close-open)*shares
         , ROI = percent(p_l/account)) %>% 
  select(cross_strat, ROI) -> bh_roi

# all together now
algo_roi %>% 
  rbind(bh_roi) %>% 
  as.data.frame() -> all_roi_train

all_roi_train
```

The good news, two EMA/SMA algos produced positive returns and the other two produce minimal negative returns! The bad news, Buy & Hold **crushed** our algos' performances. Oof. 

Re-running the backtest on the Test Set (out-of-sample) produced significantly *worse* results for our algos, while Buy & Hold continued its dominance.

```{r, echo = FALSE}
all_roi_test <- readRDS("/Users/Fennon/Desktop/My Blog/projects/ema_sma_cross/all_roi_test.rds")
all_roi_test <- as.data.frame(all_roi_test)
```

```{r, echo = FALSE}
all_roi_test
```

Damn, who would have thought Buy & Hold would be so hard to beat? /s

